---
title: "State space vector autoregressions with mvgam"
output: html_document
date: "2025-02-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mvgam)           # Fit, interrogate and forecast DGAMs
library(tidyverse)       # Tidy and flexible data manipulation
library(ggplot2)         # Flexible plotting
library(tidybayes)       # Graceful plotting of Bayesian posterior estimates
library(farver)          # Colour space manipulations

theme_set(theme_classic(base_size = 15,
                        base_family = 'serif'))
myhist = function(...){
  geom_histogram(col = 'white',
                 fill = '#B97C7C', ...)
}
hist_theme = function(){
  theme(axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank())
}
```

## 

This is a tutorial from the vignette in `mvgam` on [state-space vector autoregressions](https://ecogambler.netlify.app/blog/vector-autoregressions/). It demonstrates how the `mvgam` package and fit a state-space VAR model with a non-Gaussian observation model.

Load the data. This is from a set of annual American kestrel abundance time seres. These contain adjusted annual counts of the species taken in three regions of Britich Columbia, Canada. As formatted there are 40 years of data with a column for each of the regions.

```{r}
load(url('https://github.com/atsa-es/MARSS/raw/master/data/kestrel.rda'))
head(kestrel)

# Put the data in long format 
model_data <- kestrel |>
  as.data.frame() |>
  tidyr::pivot_longer(!Year, 
                      names_to = "region",
                      values_to = "log_count") |>
  mutate(adj_count =exp(log_count),
         series = as.factor(region),
         time = Year)
head(model_data)

plot_mvgam_series(data = model_data, 
                  y = "adj_count",
                  series = "all")
plot_mvgam_series(data = model_data,
                  y = 'adj_count',
                  series = 1)


```

The outcome variable only takes non-negative real numbers

## Fit a VAR(1) model

We will use the `trend_formula()` argument to specify we want a state-space model, and we can use the `trend_model()` to specify that we want the latent process to evolve as Vector Autoregression of order 1. This takes advantage of developments in the field that enforce stationarity through a principled prior on the autoregressive coefficients. Because magnitude of average coutns varies across regions, we will need to include region level intercepts in the process model. And since the response variable is non-negative real variables, we can choose to use a Gamma distribution for the observation model.

First look at the default prior distributions for the parameters in the model

```{r}
def_priors <- get_mvgam_priors(adj_count ~ -1,
                               trend_formula = ~region,
                               trend_model = VAR(cor = TRUE),
                               data = model_data,
                               family = Gamma())

```

We will want to adjust these priors from the defaults based on domain knowledge.

```{r}

var_mod <- mvgam(
  # Observation formula, empty to only consider the Gamma observation process
  formula = adj_count ~ -1,
  
  # Process model formula that includes regional intercepts
  trend_formula = ~ region,
  
  # A VAR(1) dynamic process with fully parameterized covariance matrix Sigma
  trend_model = VAR(cor = TRUE),
  
  priors = c(prior(std_normal(), class = Intercept_trend),
             prior(std_normal(), class = b),
             prior(exponential(2.5), class = sigma)),
  
  # The time series data in `long` format
  data = model_data,
  
  # A gamma observation family 
  family = Gamma(),
  
  # Forcing all three series to share the same Gamma shape parameter
  share_obs_params = TRUE,
  
  #Stan control arguments
  adapt_delta = 0.95,
  burnin = 1000,
  samples = 1000,
  silent = 2
)

```

```{r}
summary(var_mod)


```

Plot some things, in particular, we want to look at the process and observation scale parameters (`sigma` and `shape` in the model's MCMC output)

```{r}
conditional_effects(var_mod)
mcmc_plot(var_mod,
          variable = 'sigma',
          regex = TRUE,
          type = 'trace')
mcmc_plot(var_mod,
          variable = 'shape',
          type = 'trace')
pairs(var_mod,
      variable = c('shape', 'sigma'),
      regex = TRUE,
      off_diag_args = list(size = 1,
                           alpha = 0.5))
```
Posterior predictive checks, integrating over all the stable temporal states for
each series (if there were no observations to inform the states of the latent 
VAR(1) process)
```{r}
pp_check(var_mod,
         type = "dens_overlay_grouped",
         group = "region",
         ndraws = 50)
```

Conditional posterior predictive checks using the actual latent states
estimated when the model condition on observed data
```{r}
ppc(var_mod,
    series = 1,
    type = 'density')
```